LHC abort gap monitor electronics upgrade

### 2024-05-06 paper.tex
Update on Overleaf.

The system architecture is illustrated in
Fig.~\ref{fig:bsra_chain}, with each part labelled accordingly.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_chain.pdf}
    \caption{BSRA Chain.}
    \label{fig:bsra_chain}
\end{figure} 
The drawing depicts the layout of the LHC, with beam 1 represented
in red and beam 2 in blue. It includes all four experiments (Alice, Atlas,
LHCb, and CMS). The AGM is located at Point 4 (indicated by the yellow
square), next to the CMS experiment. The Timing, Trigger, and Control system
(TTC) is shown as a purple rectangle in the center of the LHC ring. This system
receives RF clocking and broadcasts it into the network (TTC
distribution) through an optical fiber~\cite{ttc_distribution}. The AGM extracts bunch clock 
(corresponding to the position of bunches) and turn clock (marking
the start of a~turn) from the LHC timing block messages.\\
Beam intensity is measured using the synchrotron light (SL) radiation, as indicated by the red line directed towards the Multi-channel photomultiplier
(MCP-PMT) in the optical front-end (OF). The beam traverses first Neutral Density optical filters, which increases the system dynamic range. Subsequently, the signal is
transmitted over a \SI{100}{m}-long cable. To compensate for losses in the cable,
the signal is amplified beforehand. Then it proceeds to the Data Acquisition
(DAQ) system, which measures \SI{6.4}{\micro\second} window and integrates the
intervals of \SI{100}{\nano\second}. This creates set of 64 values published to
the SW every \SI{10}{\micro\second}.

\subsection{Working principle}
The measurement period is determined by the MCP-PMT used, which is the gated
HAMAMATSU R5916U-50 with a maximum duty cycle of \SI{1}{\%}. With a gate pulse duration
of \SI{3}{\micro\second} and one turn cycle length of \SI{89}{\micro\second},
measurements are taken every $4^{\text{th}}$ clock cycle (\SI{356}{\micro\second}),
reducing the duty cycle to \SI{0.84}{\%}.\\
The system captures 64 values with a granularity of \SI{100}{\nano\second}. The
MCP-PMT gate duration is \SI{3}{\micro\second} providing an additional degree of
freedom in the system control. Each \SI{100}{\nano\second} value is averaged in an
extra Exponential Moving Average (EMA) and provided to the software every \SI{100}{\micro\second}. This
time is sufficient to react to changes in AG. The population can change from
1.6 to \SI{25}{\second} due to different
effects~\cite{high_sensitivity_measurement}.

\subsection{Current solution}
The current setup employs a DAB64x card with an FPGA Altera Stratix EP1S40 for
data processing. The platform features two FMC connectors, each hosting an IBMS
module integrating voltage input via the lhcb2002b ASIC~\cite{lhcb_asic}. To
maximize ADC resolution, the signal undergoes amplification using a HAMAMATSU
C5594 amplifier with a static gain of \SI{36}{dBV}, a bandwidth of
\SI{1.5}{GHz}, and a low cutoff frequency of \SI{50}{kHz}.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{existing_system_processing.pdf}
    \caption{Current system block diagram}
    \label{fig:current_system_processing}
\end{figure}
The system schematic is depicted in Fig.\ref{fig:current_system_processing}.
The OF remains consistent with the description provided
in the system architecture chapter. Note that filters are not represented in the
figure as they are universally applied. For optimal utilization of the \SI{40}{MHz}
ADC's full scale, the MCP-PMT signal requires a~minimum amplification of
\SI{34}{dB}, a task accomplished by the Hamamatsu C559
amplifier\cite{bsra_first_operation}.\\
The signal corresponding to a~single bunch (\SI{25}{ns})
is integrated within the IBMS module (denoted by a~blue triangle with a~capacitor in the feedback loop) and subsequently sampled by the ADC
(represented by a violet rectangle). The DAQ system integrates four samples
(indicated by a~light blue rectangle with a $\sum$ symbol) to compute the
charge over a~\SI{100}{ns} interval. This process generates 64 values, from
which two sets of 64 values are derived, encompassing the standard deviation
and mean value calculations. Both sets are then forwarded to the EMA block (depicted by a green rectangle). All collected data
are stored within the block RAM and subsequently shared with a PC via a
Wishbone to VME64 bridge. The PC operates a~server, facilitating the FESA
interface for connected clients.

\subsection{Constraints and limitations}
The present AGM exhibits artifacts in the measurements. Signals outside the gating range do not exhibit the characteristics of white noise, attributed to the lhcb2002b ASIC. The gain of the integrators lacks linearity, featuring distinct offsets and dead times. Consequently, the system requires correction for these effects at each integrator, leading to uncertainties in the outcomes and necessitating intricate calibration procedures. In addition, the amplifier bandwidth (50 kHz - 1,5 GHz) makes it necessary to implement a correction for the missing DC- 50 kHz band to correctly measure AG profiles needed during LHC commissioning phases.\\
The system relies on an outdated FPGA platform, rendering its development
software incompatible with the latest operating systems. Additionally, the
fixed amplifier configuration contributes to reduced resolution for signals of
smaller magnitudes.

\section{System Upgrade}
To address maintenance concerns, the VFC-HD carrier is employed alongside an
additional RTM board, facilitating gate driver functionality and control of the
built-in test LED. Leveraging the same carrier board equipped with a~500MSPS
14-bit ADC from previous projects meets the requirement for reducing
maintenance efforts. The updated system eliminates IBMS cards featuring
lhcb2002b integrators, transitioning the entire integration process to the
gateware.

\subsection{Low band amplifier}
The newly developed custom amplifier is designed to accommodate a wider signal
dynamic range by virtue of its variability.
% A = U^2/(i^2 * R^2) => A_dB = 20*log10(U/(i*R))
Figure \ref{fig:amp_schematics} illustrates the block diagram of the amplifier.
All configurations are managed through an~UART interface, facilitating
adjustments to the variable current-to-voltage conversion factor ranging from
\SI{5.8}{mV/\micro A} to \SI{1.1}{V/\micro A}. This corresponds to a~voltage
gain range from \SI{41}{dB} to \SI{87}{dB}. The AGM ADC does not exhibit a~zero
offset, allowing for signal shifting within the amplifier prior to
post-processing.\\
Additionally, the amplifier incorporates an alarm signal generator designed to
deactivate the gate, when the incident light intensity exceeds safe levels,
potentially risking damage to the multi-channel plate within the PMT. The alarm
signal is transmitted to the gateware for processing and management. Notably,
the alarm feature includes a~variable threshold setting.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{amp_schematics.pdf}
    \caption{New AGM custom amplifier.}
    \label{fig:amp_schematics}
\end{figure}
The amplifier design includes three switches. The first switch can configure
the pre-integrator to filter out spikes resulting from dark counts of the PMT.
The second switch enables selection between a~manual or a~gate-driven integration
of the acquisition window. The third switch activates low-pass filtering with a~cutoff frequency of \SI{40}{MHz}.

>>>>

The system architecture is illustrated in
Fig.~\ref{fig:bsra_chain}, with each part labelled accordingly.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_chain.pdf}
    \caption{BSRA Chain.}
    \label{fig:bsra_chain}
\end{figure} 
The drawing depicts the layout of the LHC, with beam 1 represented
in red and beam 2 in blue. It includes all four experiments (Alice, Atlas,
LHCb, and CMS). The AGM is located at Point 4 (indicated by the yellow
square), next to the CMS experiment. The Timing, Trigger, and Control system
(TTC) is shown as a purple rectangle in the center of the LHC ring. This system
receives RF clocking and broadcasts it into the network (TTC
distribution) through an optical fiber~\cite{ttc_distribution}. The AGM extracts bunch clock 
(corresponding to the position of bunches) and turn clock (marking
the start of a~turn) from the LHC timing block messages.\\
Beam intensity is measured using the synchrotron light (SL) radiation, as indicated by the red line directed towards the Multi-channel photomultiplier
(MCP-PMT) in the optical front-end (OF). The beam traverses first Neutral Density optical filters, which increase the system dynamic range. Subsequently, the signal is
transmitted over a \SI{100}{m}-long cable. To compensate for losses in the cable,
the signal is amplified beforehand. Then it proceeds to the Data Acquisition
(DAQ) system, which integrates the incoming signal in \SI{100}{\nano\second} intervals within a \SI{6.4}{\micro\second}-long acquisition window. This creates a set of 64 values published to
the SW every \SI{10}{\micro\second}.

\subsection{Working principle}
The measurement period is determined by the MCP-PMT used, which is the gated
HAMAMATSU R5916U-50 with a maximum duty cycle of \SI{1}{\%}. With a gate pulse duration
of \SI{3}{\micro\second} and one turn cycle length of \SI{89}{\micro\second},
measurements are taken every $4^{\text{th}}$ clock cycle (\SI{356}{\micro\second}),
reducing the duty cycle to \SI{0.84}{\%}.\\
The system captures 64 values with a granularity of \SI{100}{\nano\second}. The
MCP-PMT gate duration is \SI{3}{\micro\second} providing an additional degree of
freedom in the system control. Each \SI{100}{\nano\second} value is averaged in an
 Exponential Moving Average (EMA) and provided to the software every \SI{100}{\micro\second}. This
time is sufficient to react to changes in AG. The population can change from
1.6 to \SI{25}{\second} due to different
effects~\cite{high_sensitivity_measurement}.

\subsection{Current solution}
The current setup employs a DAB64x card with an FPGA Altera Stratix EP1S40 for
data processing. The platform features two FMC connectors, each hosting an IBMS
module, integrating voltage input via the lhcb2002b ASIC~\cite{lhcb_asic}. To
maximize ADC resolution, the signal undergoes amplification using a HAMAMATSU
C5594 amplifier with a static gain of \SI{36}{dBV}, a bandwidth of
\SI{1.5}{GHz}, and a low cutoff frequency of \SI{50}{kHz}.

\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{existing_system_processing.pdf}
    \caption{Current system block diagram}
    \label{fig:current_system_processing}
\end{figure}

The system schematic is depicted in Fig.~\ref{fig:current_system_processing}.
The OF remains consistent with the description provided
in the system architecture chapter. Note that filters are not represented in the
figure as they are universally applied. For optimal utilization of the \SI{40}{MHz}
ADC's full scale, the MCP-PMT signal requires a~minimum amplification of
\SI{34}{dB}, a task accomplished by the Hamamatsu C559
amplifier\cite{bsra_first_operation}.\\
The signal corresponding to a~single bunch (\SI{25}{ns})
is integrated within the IBMS module (denoted by a~blue triangle with a~capacitor in the feedback loop) and subsequently sampled by the ADC
(represented by a violet rectangle). The DAQ system integrates four samples
(indicated by a~light blue rectangle with a $\sum$ symbol) to compute the
charge over a~\SI{100}{ns} interval. This process generates 64 values, from
which two sets of 64 values are derived, encompassing the standard deviation
and mean value calculations. Both sets are then forwarded to the EMA block (depicted by a green rectangle). All collected data
are stored within the block RAM and subsequently shared with a PC via a
Wishbone to VME64 bridge. The PC operates a~server, facilitating the FESA
interface for connected clients.

\subsection{Constraints and limitations}
The present AGM exhibits artifacts in the measurements. Signals outside the gating range do not exhibit the characteristics of white noise, attributed to the lhcb2002b ASIC. The gain of the integrators lacks linearity, featuring distinct offsets and dead times. Consequently, the system requires correction for these effects for each integrator, leading to increased uncertainty of the measurement, necessitating intricate calibration procedures. In addition, the amplifier bandwidth (50 kHz - 1,5 GHz) makes it necessary to implement a correction for the missing DC- 50 kHz band to correctly measure AG profiles needed during LHC commissioning phases.\\
The system relies on an outdated FPGA platform, rendering its development
software incompatible with the latest operating systems. Additionally, the fixed amplifier configuration contributes to reduced resolution for signals of smaller magnitudes.

\section{System Upgrade}
To address the long-term maintenance concerns, the original system described above had to be completely refurbished. While the optical path was kept untouched, the DAQ and the front-end amplifier were replaced by a modern solution. Original analogue integrators were replaced by a 500MSPS 14-bit FMC-form factor sampler. The original amplifier was replaced by a DC-coupled amplifier, allowing for dynamic change of the gain. The FMC ADC module is installed into an in-house built FPGA carrier (VFC-HD) featuring 2$\times 8$GBits DDR3 memories and large Arria V FPGA.  Such solution eliminates completely the need for analogue integrators as all the signal processing can be implemented into the GateWare.

\subsection{Low band amplifier}

Figure \ref{fig:amp_schematics} illustrates the block diagram of the new amplifier. The amplifier was designed to be highly configurable.
% A = U^2/(i^2 * R^2) => A_dB = 20*log10(U/(i*R))

The configuration is done through an~UART interface, facilitating
setting of a variable current-to-voltage conversion factor. The conversion factor can vary from
\SI{5.8}{mV/\micro A} to \SI{1.1}{V/\micro A}, corresponding to a~voltage gain ranging from \SI{41}{dB} to \SI{87}{dB} in a $50\Omega$ system. The amplifier also implements a programmatic DC-offset compensation, which can be used in the future to extend the signal-to-noise ratio of the measurement. Further, the amplifier provides an alarm signal designed to
deactivate the PMT gate, when the incident light intensity exceeds safe levels. The alarm threshold is as well settable through UART. It is stored into the amplifier's non-volatile memory to be ready at the amplifier power-up. The alarm is transmitted to the gateware for processing and management.

\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{amp_schematics.pdf}
    \caption{New AGM custom amplifier.}
    \label{fig:amp_schematics}
\end{figure}

The amplifier design includes three switches. The first switch can configure
the pre-integrator to filter out spikes resulting from dark counts of the PMT.
The second switch enables selection between a~manual or a~gate-driven integration
of the acquisition window. The third switch activates low-pass filtering with a~cutoff frequency of \SI{40}{MHz}.
